---
title: "Allele Specific Copy Number Inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Allele Specific Copy Number Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning=FALSE, message=FALSE, 
  comment = "#>"
)
```

```{r setup}
library(signals)
```

## Background
This vignette illustrates how to perform allele specific copy number inference with DLP+ data. However the method should be generally applicable to any single cell copy number method. The input requirements are discussed below.

## Data
The data needed to perform the allele specific copy number inference are binned copy number data and cell specific haplotype calls. The copy number data and the haplotype data should use the same bins.

A small test data set of 250 cells is provided with the package, we'll load this data and inspect these 2 dataframe.

```{r}
data(haplotypes)
data(CNbins)
```

```{r}
head(CNbins)
```

The haplotype data comes in long format, but the package requires data in wide format so we will use the helper function `format_haplotypes_dlp` to convert the data to the correct format, this function will also remove any bins present in the haplotypes data frame that is not present in the CNbins data frame.
```{r}
haplotypes <- format_haplotypes_dlp(haplotypes, CNbins)
head(haplotypes)
```

## Inference

The output of `signals` is allele specific states per bin. `signals` has 2 options to perform this inference, the first, `callAlleleSpecificCN`, infers allele specific states using mirrored B allele frequencies as is typically done in bulk whole genome sequencing. This infers states of the form `A|B` where `B < A`. The second option, `callHaplotypeSpecificCN` takes into account the possibility that different alleles may be lost or gained in different cells. In order to identify these events, `signals` clusters cells and infers the haplotype phase based on regions within clusters that have large allelic imbalance. The output for this method therefore allows for `B>A`.

### `callAlleleSpecificCN`

```{r, message = FALSE}
ascn <- callAlleleSpecificCN(CNbins, haplotypes)
```

As a sanity check we can plot the B allele frequency (BAF) as a function of inferred state. We would expect to see a distribution around the expected value which is exactly what we see below.

```{r, fig.show='hold', fig.height=4 , fig.width=10}
plotBAFperstate(ascn, maxstate = 10)
```

Now we can plot the copy number heatmaps. First we'll plot the heatmap based on copy number states. We can provide the `plotHeatmap` function with a tree in newick format to plot an ordered heatmap and/or a dataframe with cluster assignments. By default the ```plotHeatmap``` function does some clustering using hdbscan if neither of these is provided. 

```{r}
cl <- umap_clustering(ascn$data)
```

```{r, fig.show='hold', fig.height=4 , fig.width=10}
plotHeatmap(ascn, 
            plotcol = "state", 
            clusters = cl$clustering,
            tree = cl$tree, 
            reorderclusters = TRUE, #order according to tree
            plottree = FALSE) #do not plot the tree but keep the ordering )
```

Now we can plot some heatmaps to visualize the allele specific state. This heatmap plots the minor copy number state.
```{r, fig.show='hold', fig.height=4 , fig.width=10}
plotHeatmap(ascn, plotcol = "Min",
            clusters = cl$clustering,
            tree = cl$tree, 
            reorderclusters = TRUE, #order according to tree
            plottree = FALSE) #do not plot the tree but keep the ordering )
```

Discrete representation of allelic imbalance:
```{r, fig.show='hold', fig.height=4 , fig.width=10}
plotHeatmap(ascn, 
            plotcol = "state_phase", 
            clusters = cl$clustering,
            tree = cl$tree, 
            reorderclusters = TRUE, #order according to tree
            plottree = FALSE) #do not plot the tree but keep the ordering )
```

There are also some function to plot per cell BAF and state plots. You can specify a cell, or the default will take the first cell in the data frame.
```{r, fig.show='hold', fig.height=5, fig.width=10}
plotCNprofileBAF(ascn)
```

Here we'll plot an equivalent plot merging data across cell clusters using the utility function `consensyscopynumber`. Here, the `cell_id` column becomes the 
```{r, fig.show='hold', fig.height=5 , fig.width=10}

consensus_clusters <- consensuscopynumber(ascn$data, cl = cl$clustering)

plotCNprofileBAF(consensus_clusters, cellid = "A")
```

The object returned by the inference function is of type `ascn`, if you want to pull out the data to use in your own scripts you can just do `ascn$data`. This will return a `data.frame` similar to the input data frames but with additional columns specifying the allele specific states.

### `callHaplotypeSpecificCN`
Now we'll do the analysis that phases haplotypes across cells.
```{r, message = FALSE}
hscn <- callHaplotypeSpecificCN(CNbins, haplotypes)
```

We'll now generate the same plots as before. We can see in the BAF per state plot that we now observe states with BAF > 0.5 meaning that we see different alleles gained or lost in different cells.

```{r, fig.show='hold', fig.height=4 , fig.width=10}
plotBAFperstate(hscn, maxstate = 10)
```

In the allelic imbalance heatmap we can see that some cells have gained the "B" allele and others the "A" allele resulting in imbalance spanning from 0 to 1.

In the discrete version of this plot we see similar patterns. 
```{r, fig.show='hold', fig.height=4 , fig.width=10}
plotHeatmap(hscn, plotcol = "state_phase",
            clusters = cl$clustering,
            tree = cl$tree, 
            reorderclusters = TRUE, #order according to tree
            plottree = FALSE) #do not plot the tree but keep the ordering 
```

```{r, fig.show='hold', fig.height=4 , fig.width=10}
plotHeatmap(hscn, plotcol = "Min",
            clusters = cl$clustering,
            tree = cl$tree, 
            reorderclusters = TRUE, #order according to tree
            plottree = FALSE) #do not plot the tree but keep the ordering 
```

```{r, fig.show='hold', fig.height=5, fig.width=10}
plotCNprofileBAF(hscn)
```

```{r, fig.show='hold', fig.height=5 , fig.width=10}
consensus_clusters <- consensuscopynumber(hscn$data, cl = cl$clustering)

plotCNprofileBAF(consensus_clusters, cellid = "B")
```


An alternative plotting style is to plot the copy number of the individual copy number.
```{r, fig.show='hold', fig.height=5 , fig.width=10}
plotCNprofileBAF(consensus_clusters, cellid = "C", homolog = TRUE)
```
