---
title: "Beta-Binomial model for schnapps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beta-Binomial model for schnapps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning=FALSE, message=FALSE, 
  comment = "#>"
)
```

```{r setup}
library(schnapps)
```

## Background
This vignette demonstrates how to use a beta-binomial model instead of a binomial model for the HMM emission model.

## Inference
```{r}
data("haplotypes")
data("CNbins")
haplotypes <- format_haplotypes_dlp(haplotypes, CNbins)
hscn <- callHaplotypeSpecificCN(CNbins, haplotypes, likelihood = "binomial")

print(hscn)

hscn_bb <- callHaplotypeSpecificCN(CNbins, haplotypes, likelihood = "betabinomial")

print(hscn_bb)

```
We see that the model inferred a modest degree of overdispersion in the data, but that this was siginificantly different from 0 dispersion (Tarones Z-score is high).

Below I'll plot the heatmaps for the two output, which in this case are very similar.

### Binomial model
```{r, fig.show='hold', fig.height=4 , fig.width=10}
plotHeatmap(hscn, plotcol = "state_BAF", plottree = FALSE, spacer_cols = 15)
```

### Beta-Binomial model
```{r, fig.show='hold', fig.height=4 , fig.width=10}
plotHeatmap(hscn_bb, plotcol = "state_BAF", plottree = FALSE, spacer_cols = 15)
```

### QC plot
To interrogate the difference between the binomial and beta-binomial models we can plot the BAF and overlay the two models.

```{r, fig.show='hold', fig.height=5, fig.width=10}
plotBBfit(hscn_bb)
```

We can check how similar the results are by comparing the two dataframes. We find for this data that the results are very similar.
```{r}
print(dim(hscn$data))
print(dim(hscn_bb$data))
all.equal(orderdf(hscn$data), orderdf(hscn_bb$data))
```

Another thing we can check is the total number of segments identified by the HMM. Using the beta-binomial model should reduce the influence of any noisy regions and thus we might expect fewer segments. We do observe fewer segments in the beta-binomial model.

```{r}
segs <- create_segments(hscn$data, field = "state_AS_phased")

segs_bb <- create_segments(hscn_bb$data, field = "state_AS_phased")

print(dim(segs))
print(dim(segs_bb))
```


## Summary
With this dataset, the difference between the nbinomial model and beta-binomial model is minimal but produces a statistically superior fit, in lower quality datasets with for example lower depth of coverage the beta-binomial model may may provide more noticeable differences.

