---
title: "Structural variant visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structural variant visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.width = 9,
  fig.height = 3.2
)
```

```{r setup}
library(signals)
library(dplyr)
library(ggplot2)
```

`plotCNprofile` can draw structural variants on top of a copy number profile.
This vignette covers the data format it expects, then works from a single SV up
to whole-genome views with many.

The example data is the OV2295-derived cell line SA921: `CNbins` holds the copy
number, `SVs` holds destruct breakpoint calls for the same sample.

```{r}
data(CNbins)
data(SVs)

# CNbins bundles several samples; keep the one the SVs were called on
cn <- dplyr::filter(CNbins, grepl("^SA921", cell_id))
cell <- unique(cn$cell_id)[1]
```

## SV data format

`SV` is a data frame with one row per breakpoint. Two columns describe each
breakend, plus the read support:

```{r}
head(SVs, 3)
```

| Column | Required | Notes |
|---|---|---|
| `chromosome_1`, `position_1`, `strand_1` | yes | first breakend |
| `chromosome_2`, `position_2`, `strand_2` | yes | second breakend |
| `read_count` | for `"lines_and_arcs"` | supporting reads |
| `type`, `rearrangement_type` | optional | used by the `"curves"` style |

Strands must be the characters `"+"` and `"-"`. Callers that store them as
logicals need converting first, otherwise the arcs are silently dropped:

```{r, eval = FALSE}
SVs <- dplyr::mutate(SVs,
                     strand_1 = ifelse(strand_1, "+", "-"),
                     strand_2 = ifelse(strand_2, "+", "-"))
```

The strand pair determines how each SV is coloured. `classify_sv_orientation`
is what `plotCNprofile` uses internally:

```{r}
SVs %>%
  mutate(orientation = signals:::classify_sv_orientation(.)) %>%
  count(orientation, type)
```

| Orientation | Meaning | Copy number effect |
|---|---|---|
| `-+` | tandem duplication | gain |
| `+-` | deletion-type | loss |
| `++` / `--`, short range | foldback inversion | gain |
| `++` / `--`, long range | inversion | neutral |
| `Translocation` | inter-chromosomal | context dependent |

Positions do not have to be ordered: for intra-chromosomal SVs where
`position_1 > position_2`, `plotCNprofile` flips the breakends and their strands
so the orientation stays correct.

## One SV

Start with a single deletion-type SV on chromosome 5. Its two breakends bracket
a region where copy number drops from 4 to 2, which is what you want to see —
the SV and the copy number change are the same event.

```{r}
sv_chr5 <- SVs %>%
  filter(chromosome_1 == "5", chromosome_2 == "5",
         position_1 > 126e6, position_1 < 127e6)
sv_chr5

plotCNprofile(cn, cellid = cell, chrfilt = "5", maxCN = 10,
              SV = sv_chr5, sv_style = "lines_and_arcs")
```

The vertical line marks each breakend and the arc joins them. Height encodes
`read_count` against the secondary axis on the right.

## Two styles

`sv_style` controls how SVs are drawn. `"curves"` (the default) draws bezier
curves coloured by `rearrangement_type`; `"lines_and_arcs"` draws the lines and
arcs above, coloured by strand orientation. `"both"` draws each.

```{r}
plotCNprofile(cn, cellid = cell, chrfilt = "5", maxCN = 10,
              SV = sv_chr5, sv_style = "curves")
```

## Many SVs

With a whole chromosome's worth of SVs the picture degrades quickly. Chromosome
1 is the most rearranged in this sample:

```{r}
sv_chr1 <- SVs %>% filter(chromosome_1 == "1", chromosome_2 == "1")
nrow(sv_chr1)

plotCNprofile(cn, cellid = cell, chrfilt = "1", maxCN = 10,
              SV = sv_chr1, sv_style = "lines_and_arcs")
```

Two problems: the arcs sit on top of the copy number points, and because each
arc's height comes from its read count, every arc peaks somewhere different.

## Moving the arcs into their own band

`sv_arcs_above = TRUE` draws the arcs in a band above the copy number panel
instead. Read count no longer affects the geometry, so the secondary axis is
dropped.

```{r}
plotCNprofile(cn, cellid = cell, chrfilt = "1", maxCN = 10,
              SV = sv_chr1, sv_style = "lines_and_arcs",
              sv_arcs_above = TRUE)
```

The copy number track is now unobstructed. `sv_band_frac` sets how much of the
panel the band takes:

```{r}
plotCNprofile(cn, cellid = cell, chrfilt = "1", maxCN = 10,
              SV = sv_chr1, sv_style = "lines_and_arcs",
              sv_arcs_above = TRUE, sv_band_frac = 0.4)
```

### Arc height

By default the apex of each arc scales with the distance it spans, so local
rearrangements stay low and long-range ones arch over them. `sv_arc_scale =
"fixed"` gives every arc the same height instead, which makes short SVs into
narrow spikes:

```{r}
plotCNprofile(cn, cellid = cell, chrfilt = "1", maxCN = 10,
              SV = sv_chr1, sv_style = "lines_and_arcs",
              sv_arcs_above = TRUE, sv_arc_scale = "fixed")
```

Two floors keep small SVs visible. `sv_arc_min_width` matters most: an SV whose
breakends fall in the same bin has zero width and would not render at all —
which at 500kb bins is every foldback. `sv_arc_min_height` sets the apex of the
shortest arcs.

```{r}
plotCNprofile(cn, cellid = cell, chrfilt = "1", maxCN = 10,
              SV = sv_chr1, sv_style = "lines_and_arcs",
              sv_arcs_above = TRUE,
              sv_arc_min_height = 0.5, sv_arc_min_width = 0.01)
```

### Splitting the band by copy number effect

`sv_arc_side` puts the baseline in the middle of the band and splits arcs either
side of it. With `"cn_effect"`, gain-associated junctions (tandem duplications,
foldbacks, translocations) arch up and loss-associated or copy-number-neutral
ones hang down, so an arc's side predicts the copy number step beneath it.

```{r}
plotCNprofile(cn, cellid = cell, chrfilt = "1", maxCN = 10,
              SV = sv_chr1, sv_style = "lines_and_arcs",
              sv_arcs_above = TRUE, sv_band_frac = 0.4,
              sv_arc_side = "cn_effect")
```

`"translocation"` and `"foldback"` are the other built-in rules. You can also
pass the name of a column in `SV` holding `"up"`/`"down"` to split on anything
you like:

```{r}
sv_chr1_side <- sv_chr1 %>%
  mutate(side = ifelse(read_count >= 30, "up", "down"))

plotCNprofile(cn, cellid = cell, chrfilt = "1", maxCN = 10,
              SV = sv_chr1_side, sv_style = "lines_and_arcs",
              sv_arcs_above = TRUE, sv_band_frac = 0.4,
              sv_arc_side = "side")
```

### Breakpoint lines

The vertical lines run from 0 up to the arc baseline and are drawn behind the
copy number points. At high SV density they can dominate; `sv_show_lines =
FALSE` leaves the arcs alone.

```{r}
plotCNprofile(cn, cellid = cell, chrfilt = "1", maxCN = 10,
              SV = sv_chr1, sv_style = "lines_and_arcs",
              sv_arcs_above = TRUE, sv_show_lines = FALSE)
```

`sv_line_alpha` controls the lines independently of `sv_arc_alpha`, which is
useful because the lines sit behind the data and can take more weight than the
arcs.

## Translocations

A translocation is only drawn when **both** breakends are inside the plotted
region, so inter-chromosomal SVs need both chromosomes in view. Passing several
chromosomes to `chrfilt` works, but `regions` is usually better: it crops to the
interesting windows and keeps the arcs between them.

```{r, fig.height = 3.6}
regions <- data.frame(chr = c("17", "2"), start = c(60, 5), end = c(70, 50))

sv_1702 <- SVs %>%
  filter((chromosome_1 == "17" & chromosome_2 == "2") |
         (chromosome_1 == "2"  & chromosome_2 == "17"))

plotCNprofile(cn, cellid = cell, regions = regions, maxCN = 10,
              SV = sv_1702, sv_style = "lines_and_arcs",
              sv_arcs_above = TRUE, sv_band_frac = 0.35)
```

`show_chrbreaks = FALSE` removes the grey lines that bracket each region, and
`ybreaks` overrides the y axis breaks — useful in short panels, where the
defaults collide.

```{r, fig.height = 3.6}
plotCNprofile(cn, cellid = cell, regions = regions, maxCN = 10,
              SV = sv_1702, sv_style = "lines_and_arcs",
              sv_arcs_above = TRUE, sv_band_frac = 0.35,
              show_chrbreaks = FALSE, ybreaks = c(0, 5, 10))
```

## Whole genome

Everything works genome-wide, though at this scale only the broadest structure
is legible.

```{r, fig.height = 3.6}
plotCNprofile(cn, cellid = cell, maxCN = 10,
              SV = SVs, sv_style = "lines_and_arcs",
              sv_arcs_above = TRUE, sv_band_frac = 0.3,
              sv_show_lines = FALSE, sv_arc_alpha = 0.6)
```

## Legend

Orientation colours are off by default. `show_sv_legend = TRUE` adds them to the
plot, or `get_sv_lines_and_arcs_legend()` returns a standalone legend for
assembling figures by hand.

```{r, fig.height = 2}
cowplot::plot_grid(get_sv_lines_and_arcs_legend(direction = "horizontal"))
```
